## 3.5.1 Changing Projections 

Download a geopackage for vector of Australia.
Seeing what is inside it using st_layers*()

```{r}

#packages
library(sf)
library(here)

install.packages(c("sf", "tmap", "tmaptools", "RSQLite", "tidyverse"))
install.packages("dplyr")
install.packages("readr")
install.packages("tidyverse")
library("tidyverse")
library("readr")
library("dplyr")

#spatial data packages

install.packages("maptools")
install.packages(c("classInt", "tmap"))

# might also need these ones
install.packages(c("RColorBrewer", "sp", "rgeos", 
                   "tmaptools", "sf", "downloader", "rgdal", 
                   "geojsonio"))

library(maptools)
library(RColorBrewer)
library(classInt)
library(sp)
library(rgeos)
library(tmap)
library(tmaptools)
library(sf)
library(rgdal)
library(geojsonio)



#from practical

library(sf)
library(here)
st_layers(here("prac3_data", "gadm36_AUS.gpkg"))
```
Now read in the GeoPackage layer for the whole of Australia 

```{r}
library(sf)
Ausoutline <- st_read(here("prac3_data", "gadm36_AUS.gpkg"), 
                      layer='gadm36_AUS_0')
                      
```
we can check the coordinate reference systems of sf or sp objects using the print function:

```{r}
print(Ausoutline)
```
here, the coordinate reference system is in Geodetic CRS: WGS 84. Geodetic means geographic. 

Another way of identifying the CRS is to find the proj4 string. A proj4 string is meant to be a compact way of identifying a coordinate reference system.
```{r}
library(sf)
st_crs(Ausoutline)$proj4string
```
if our data didn't have a spatial reference system, we could have just set it using st_set_crs()

this works by erasing what was there before and replacing it with the crs you want. 

however it is only useful if there is no CRS when you load the data.

```{r}
Ausoutline <- Ausoutline %>%
  st_set_crs(., 4326)

# or more concisely

Ausoutline <- st_read(here("prac3_data", "gadm36_AUS.gpkg"), 
                      layer='gadm36_AUS_0') %>% 
  st_set_crs(4326)

```
## Reprojecting your spatial data

For generating maps in packages like leaflet, maps need to be in WGS84 rather than a projected flat reference system.

But projected reference systems are useful for things such as measuring the distance of a line object, or the distance between two polygons. 


Once your data has a coordinate system, we can re-project or transform to anything we like. 
Using st_transform

3112 is the EPSG code
```{r}
AusoutlinePROJECTED <- Ausoutline %>%
  st_transform(.,3112)

print(AusoutlinePROJECTED)
```
you might encounter an sp object from the sp package. in this case i'd advise just transforming the sp object to sf and changing the projection

```{r}
#from sf to sp
AusoutlineSP <- Ausoutline %>%
  as(., "Spatial")

#From sp to sf
AusoutlineSF <- AusoutlineSP %>%
  st_as_sf()
```

## 3.5.2 WorldClim Data

loading in raster data from worldclim for precipitation at 5m resolution
```{r}
library(raster)
library(terra)

jan <- terra::rast(here("prac3_data", "wc2.1_5m_prec", "wc2.1_5m_prec_01.tif"))
  
jan

plot(jan)
```
#### reprojecting 

to reproject a raster, the whole grid must be recomputed, and the attributes then reestimated to a new grid. 

we have to use the project() from the terra package. 

```{r}
# set the proj 4 to a new object

pr1 <- terra::project(jan, "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")

#or....

newproj<-"ESRI:54009"
# get the jan raster and give it the new proj4
pr1 <- jan %>%
  terra::project(., newproj)
plot(pr1)
```

plotting back to WGS84

```{r}

pr1 <- pr1 %>%
  terra::project(., "EPSG:4326")
plot(pr1)

```
## 3.5.3 Data loading

1. list all the files stored within our directory with dir_info() from the fs package, then use dplyr in conjunction with str_detect() from stringr to search for filenames containing tif. 

FIRST, dir_info()
```{r}
#look in our folder, find the files that end with .tif

library(fs)
dir_info("prac3_data/")

```
we can use this with dplyr to select the data we actually want. select() exists both within dplyr and raster package so make sure you use the right one.

using dplyr::select forces select from dplyr

```{r}

# pull() from dplyr is the same as $ often used to extract columns, as in the next stage the input must be file names as characters (nothing else like a column name)

library(tidyverse)
listfiles<-dir_info("prac3_data/wc2.1_5m_prec") %>%
  filter(str_detect(path, ".tif")) %>%
  dplyr::select(path)%>%
  pull()

#have a look at the file names
listfiles
```
##SpatRaster

1. then load all of the data straight into a SpatRaster. A SpatRaster is a collection of raster layers with the same spatial extent and resolution

it will stack all the raster files

```{r}
worldclimtemp <- listfiles %>%
  terra::rast()

#look at the raster stack
worldclimtemp
```
under dimensions it says there are 12 layers (nlyr). the stack has loaded 12 months of average percipitation data for us in order.

## to access single layers within a stack:

```{r}
#to access the january layer
worldclimtemp[[1]]
```
## rename our layers within the stack

```{r}
month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

names(worldclimtemp) <- month

```

## get data for just January by using our new layer name

```{r}
worldclimtemp$Jan
```
# 3.5.4 Raster Location

using a raster stack we can extract data with a single command. 
for example, lets make a df of some sample sites - australian cities/towns

```{r}

site <- c("Brisbane", "Melbourne", "Perth", "Sydney", "Broome", "Darwin", "Orange", 
          "Bunbury", "Cairns", "Adelaide", "Gold Coast", "Canberra", "Newcastle", 
          "Wollongong", "Logan City" )

lon <- c(153.03, 144.96, 115.86, 151.21, 122.23, 130.84, 149.10, 115.64, 145.77, 
         138.6, 153.43, 149.13, 151.78, 150.89, 153.12)

lat <- c(-27.47, -37.91, -31.95, -33.87, 17.96, -12.46, -33.28, -33.33, -16.92, 
         -34.93, -28, -35.28, -32.93, -34.42, -27.64)

#Put all of this information into one list 

samples <- data.frame(site, lon, lat, row.names="site")

# Extract the data from the Rasterstack for all points 

AUcitytemp<- terra::extract(worldclimtemp, samples)
```

add the city names to the rows of AUcitytemp

```{r}
AUcitytemp2 <- AUcitytemp %>%
  as_tibble()%>%
  add_column(Site = site, .before = "Jan")
```

##3.6 PART 2: Descriptive statistics

## 3.6.1 Data preparation

Let's take Perth as an example. We can subset our data either using the row name:

```{r}
Perthtemp <- AUcitytemp2 %>%
  filter(site=="Perth")
```

or the row location:

```{r}
Perthtemp <- AUcitytemp2[3,]
```

##3.6.2 Histogram

making a histogram of Perth's precipitation

the tibble stored the data as double and the base hist() function needs it as numeric

```{r}
hist(as.numeric(Perthtemp))
```
## improving the aesthetics of the histogram

```{r}
library(tidyverse)
#define where you want the breaks in the historgram
userbreak<-c(0,25,50,100,125,150,175,200,225)

# remove the ID and site columns
Perthtemp <- AUcitytemp2 %>%
  filter(site=="Perth")

t<-Perthtemp %>%
 dplyr::select(Jan:Dec)

  hist((as.numeric(t)), 
     breaks=userbreak, 
     col="red", 
     main="Histogram of Perth Temperature", 
     xlab="Temperature", 
     ylab="Frequency")
```
check the histogram information R generated

```{r}
histinfo <- as.numeric(t) %>%
  as.numeric()%>%
  hist(.)

histinfo
```
here we have:

breaks - cut off points for the bins
counts - number of cells in each bin
midpoints - middle value for each bin
density - density of data per bin

## 3.6.3 Using more data

what if we wanted to see the distribution of precipitation for the whole of Australia in Jan as opposed to just our point for perth.

check the layer by plotting the geometry

```{r}
plot(Ausoutline$geom)
```
as the .shp is quite complex (lots of points) we can simplify it with the st_simplify() function. 

here the argument dTolerance controls the level of generalisation in the units of the map

you can also set preserveTopology to TRUE or FALSE
when TRUE: polygons arent removed and holes in the polygons are retained.

```{r}
AusoutSIMPLE <- Ausoutline %>%
  st_simplify(., dTolerance = 1000) %>%
  st_geometry()%>%
  plot()
```
for best analysis, recommend using the real file to ensure you dont simplyfy a potentially important variable

## next setting our map extent - where we want to clip the data

however we need to make sure that both our layers are in the same coordinate reference system when we combine them

```{r}
print(Ausoutline)
```
```{r}
crs(worldclimtemp)
```
since the crs is the same we can continue

```{r}
Austemp <- Ausoutline %>%
  # now crop our temp data to the extent
  terra::crop(worldclimtemp,.)

# plot the output
plot(Austemp)
```
if we want to just get raster data within the outline of the shape use the mask() function which uses a cookie cutter of the raster with the vector layer:

```{r}
exactAus<-terra::mask(Austemp, Ausoutline)
```
recomputing histogram for Australia in March

```{r}
#subset using the known location of the raster
hist(exactAus[[3]], col="red", main ="March precipitation")
```
## Histogram with ggplot

we need to make our raster into a data.frame to be compatible with ggplot2, using a dataframe or tibble

```{r}

exactAusdf <- exactAus %>%
  as.data.frame()

```
plot the ggplot2

```{r}
library(ggplot2)
# set up the basic histogram
gghist <- ggplot(exactAusdf, 
                 aes(x=Mar)) + 
  geom_histogram(color="black", 
                 fill="white")+
  labs(title="Ggplot2 histogram of Australian March precipitation", 
       x="Temperature", 
       y="Frequency")
# add a vertical line to the hisogram showing mean precipitation
gghist + geom_vline(aes(xintercept=mean(Mar, 
                                        na.rm=TRUE)),
            color="blue", 
            linetype="dashed", 
            size=1)+
  theme(plot.title = element_text(hjust = 0.5))
```
## plotting multiple months of precipitation data on the same histogram

we need to put variable (month) into one column using pivot_longer(). 

we are saying select columns 1-12 and place them in a new column called Month and their values in another called Precip

```{r}
squishdata<-exactAusdf%>%
  pivot_longer(
  cols = 1:12,
  names_to = "Month",
  values_to = "Precip"
)
```

then subset the data, selecting two months using filter() from dplyr

```{r}
twomonths <- squishdata %>%
  # | = OR
  filter(., Month=="Jan" | Month=="Jun")
```

get the mean for each month we selected, using group_by() and summarise()

```{r}
meantwomonths <- twomonths %>%
  group_by(Month) %>%
  summarise(mean=mean(Precip, na.rm=TRUE))

meantwomonths
```
select colour and fill based on variable.
the intercept is the mean we calculated

```{r}
ggplot(twomonths, aes(x=Precip, color=Month, fill=Month)) +
  geom_histogram(position="identity", alpha=0.5)+
  geom_vline(data=meantwomonths, 
             aes(xintercept=mean, 
                 color=Month),
             linetype="dashed")+
  labs(title="Ggplot2 histogram of Australian Jan and Jun
       precipitation",
       x="Precipitation",
       y="Frequency")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

1. drop all the NAs with drop_na()
2. made sure that the month column has the levels spcified, which will map in descending order
3. selected a bin width of 5 and produced a faceted plot

```{r}
data_complete_cases <- squishdata %>%
  drop_na()%>% 
  mutate(Month = factor(Month, levels = c("Jan","Feb","Mar",
                                          "Apr","May","Jun",
                                          "Jul","Aug","Sep",
                                          "Oct","Nov","Dec")))

# Plot faceted histogram
ggplot(data_complete_cases, aes(x=Precip, na.rm=TRUE))+
  geom_histogram(color="black", binwidth = 10)+
  labs(title="Ggplot2 faceted histogram of Australian temperatures", 
       x="Temperature",
       y="Frequency")+
  facet_grid(Month ~ .)+
  theme(plot.title = element_text(hjust = 0.5))
```
##interactive histogram using plotly

```{r}
library(raster)
library(ggplot2)
library(stats)
library(plotly)
library(graphics)
# split the data for plotly based on month

jan <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jan")

jun <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jun")

# give axis titles
x <- list (title = "Temperature")
y <- list (title = "Frequency")

# set the bin width
xbinsno<-list(start=0, end=40, size = 2.5)

# plot the histogram calling all the variables we just set
ihist<-plot_ly(alpha = 0.6) %>%
        add_histogram(x = jan$Precip,
        xbins=xbinsno, name="January") %>%
        add_histogram(x = jun$Precip,
        xbins=xbinsno, name="June") %>% 
        layout(barmode = "overlay", xaxis=x, yaxis=y)

ihist
```
## other descriptive stats

```{r}
# mean per month
meanofall <- squishdata %>%
  group_by(Month) %>%
  summarise(mean = mean(Precip, na.rm=TRUE))

# print the top 1
head(meanofall, n=1)
```
```{r}
# standard deviation per month
sdofall <- squishdata %>%
  group_by(Month) %>%
  summarize(sd = sd(Precip, na.rm=TRUE))

# maximum per month
maxofall <- squishdata %>%
  group_by(Month) %>%
  summarize(max = max(Precip, na.rm=TRUE))

# minimum per month
minofall <- squishdata %>%
  group_by(Month) %>%
  summarize(min = min(Precip, na.rm=TRUE))

# Interquartlie range per month
IQRofall <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Precip, na.rm=TRUE))

# perhaps you want to store multiple outputs in one list..
lotsofstats <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Precip, na.rm=TRUE), 
            max=max(Precip, na.rm=T))

# or you want to know the mean (or some other stat) 
#for the whole year as opposed to each month...

meanwholeyear=squishdata %>%
  summarize(meanyear = mean(Precip, na.rm=TRUE))
```

